{\pkg}: A Python Package for Integrating Computational Models Across Languages and Scales

### 2019-01-21 papers/lang2018/ms.tex
Matts feedback

%---------------------------------------------------------------------------------
% CONTROL THINGS
\newcommand{\todo}[1]{{\color{red}{#1}}}
\newcommand{\addref}{{\color{red}(ref. needed)}}
\newcommand{\Msol}{\,M$_{\odot}$}
\newcommand{\TB}{\,Tbytes}
\newcommand{\pkg}{{\tt \todo{cis\_interface}}{}}
\newcommand{\pkgrun}{{\tt \todo{cisrun}}}
\newcommand{\A}{Appendix{ }}

@@

%---------------------------------------------------------------------------------
% ABSTRACT (145 words)
\begin{abstract}
Thousands of computational models have been created within the plant biology community and broader scientific communities in the past two decades that have the potential to be combined into complex integration networks capable of capturing more complex biological processes than possible with isolated models. However, the technological barriers introduced by differences in language and data formats has slowed this progress. We present \href{https://github.com/cropsinsilico/cis_interface}{\pkg}, a Python package for running integration networks with connections between models across languages and scales. {\pkg} coordinates parallel execution of models in Python, C, C++, and Matlab on Linux, Mac OS, and Windows operating systems, and handles communication in a number of data formats common to computational plant modeling. {\pkg} is designed to be user-friendly and can be accessed at {\tt https://github.com/cropsinsilico/cis\_interface}. Although originally developed for plant models, {\pkg} can be used to connect computational models from any domain.
\end{abstract}

@@

Consider the task of integrating a root growth model with a shoot growth model. The root growth model is written in C (see Listing \ref{lst:src:root}) and can be expressed as
\begin{equation}
R_{t+1} = R_{t} \times r_{r} \times dt + R_{t}.
\end{equation}
The inputs to the root growth model are
\begin{itemize}
	\item $R_{t}$: the root mass at time $t$
	\item $r_{r}$: the relative root growth rate
	\item $dt$: the time step
\end{itemize}
and the output of the root growth model is $R_{t+1}$, the root mass at the next time step. 
%
The shoot growth model is written in Python (see Listing \ref{lst:src:shoot}) and can be expressed as
\begin{equation}
S_{t+1} = S_{t} \times r_{s} \times dt + S_{t} - (R_{t+1} - R_{t}).
\end{equation}
The inputs to the shoot model are 
\begin{itemize}
	\item $S_{t}$: the shoot mass at time $t$
	\item $r_{s}$: the relative shoot growth rate
	\item $dt$: the time step
	\item $R_{t}$: the root mass at time $t$ 
	\item $R_{t+1}$: the root mass at time $t+1$ 
\end{itemize}
and the output of the shoot growth model is $S_{t+1}$, the shoot mass at the next time step.
%
While it is possible to run both models in isolation if all of the appropriate input variables are provided, the value of $R_{t+1}$ calculated by the root growth model could be used as input to the shoot growth model. However, because the two models are written in different languages, the two models cannot be directly integrated. To integrate the models a scientist must either "manually" integrate them by running one model and then the other or translate one of the models into the language of the other so that it can be called directly. For a tightly coupled set of models such as these where one model depends directly on the result from the other at the current time step, manual integration is very inefficient for more than a handful of time steps even when done using a script. Although translating one or both of these simple toy models would be relatively straight forward, actual plant models are much more complex and could contain thousands of lines of code that would be time consuming to translate and result in unnecessary replication of model algorithms.

We present an Open Source Python package, {\pkg}, for creating and running integration networks by connecting existing computational models written in different programming languages. {\pkg} was developed as part of the Crops \emph{in Silico} \citep{Marshall-Colon2017} initiative to build a complete crop \emph{in silico} from the level of the genes to the level of the field. {\pkg} is available on Github\footnote{{\tt https://github.com/cropsinsilico/cis\_interface}} with full documentation\footnote{{\tt https://cropsinsilico.github.io/cis\_interface/}} and can be installed from PyPI\footnote{{\tt https://pypi.org/project/cis\_interface/}} via {\tt pip install {\pkg}} or conda-forge\footnote{{\tt https://github.com/conda-forge/cis\_interface-feedstock}} via {\tt conda -c conda-forge install {\pkg}}. Although developed for the purpose of integrating plant biology models, {\pkg} can be applied to any situation requiring coordination between models in the supported languages. \S\ref{S:background} provides background information on existing efforts for cross-language model integration, \S\ref{S:methods} describes the methods used by the {\pkg} package to integrate models, \S\ref{S:example} provides a worked example of integrating the two model presented above, \S\ref{S:results} presents several tests demonstrating the performance of message passing between integrated models, and \S\ref{S:discuss} describes a few use cases for {\pkg}, 
summarizes the features and limitations of the {\pkg} package, and outlines areas of ongoing and future development.

@@

These three questions can be addressed through manual integration (e.g. running the root model, converting the output root mass into the units expected by the shoot model, and then running the shoot model). However, manual integration is 1) computationally inefficient when many iterations are necessary, such as during parameter searches or steady state convergence tests, 2) unique to every integration, requiring the production of new scripts, and 3) complex when more than three models are being integrated. Within both computational biology and the larger scientific computing community, groups have developed around different solutions to the problem of connecting models. These solutions can generally be grouped into two categories: domain specific languages and frameworks.

@@

%---------------------------------------------------------------------------------
% ORCHESTRATION
\subsection{Orchestration}\label{SS:orchestration}
%
{\pkg} is executed via a command line interface (CLI), {\pkgrun} with specification files (\S\ref{SSS:yaml}) as input. Based on the information contained in the specification files, {\pkg} dynamically establishes a network of asynchronous communication channels (\S\ref{SS:communication}), and launches the models on new processes (\S\ref{SS:model_execution}). Although written in Python, {\pkg} was written such that users need not have any knowledge of the Python language and can interact with {\pkg} solely through the CLI.
%At execution, {\pkg} launches models on new processes based on information contained in a YAML specification file (See \S\ref{SSS:yaml} below) and establishes

@@

Model YAMLs include information about the location of the model source code, the language the model is written in, how the model should be run, and any input or output variables including their data type and units. 
%
%For example, Listings \ref{lst:yml:root} and \ref{lst:yml:shoot} from \A\ref{A:yml} are the model YAMLs for the example root and shoot growth models from \S\ref{S:intro}. These model YAMLs could be used in any integration without being modified because the model inputs and outputs will not change.
%
%\rootyml
%
%\shootyml
%
Integration networks are specified by declaring the connections between models and connections are declared by pairing an output variable from one model with the input variable of another model. The {\pkg} CLI sets up the necessary communication mechanisms to then direct data from one model to the next in the specified pattern. Models can have as many input and/or output variables as is desired and connections between models are specified by references to the input/output variables associated with each model. In addition, input and output variables can also be connected to files.
%
%For example, in the connection YAML for integrating the root and shoot models (Listing \ref{lst:yml:int} from \A\ref{A:yml}), the first two connections link the {\tt root\_input} and {\tt shoot\_input} input channels from the root and shoot model YAML's to input files. The third connection links the {\tt next\_root\_mass} output variable from the root model YAML (Listing \ref{lst:yml:root}) to the {\tt next\_root\_mass} input variable from the shoot model YAML (Listing \ref{lst:yml:shoot}). The forth connection links the {\tt next\_shoot\_mass} to an output file.
%
%\intyml
%
This format allows users the flexibility to create complex integration networks and test models in isolation before running the entire integration network as will be seen in \S\ref{S:example}.

%---------------------------------------------------------------------------------
% MODEL EXECUTION
\subsubsection{Model Execution}\label{SS:model_execution}
%
{\pkg} launches each model in an integration on its own process, allowing models to complete independent operations in parallel and complete tasks more quickly. For example, the root and shoot growth models from \S\ref{S:intro} (described further in \S\ref{S:example}) require 10.28\,s and 10.40\,s respectively to run for 100 time steps in isolation with direct input/output from/to files in their native language. If these two models were manually integrated in serial via the command line, the integration would require a total of 20.68\,s. However, the same integration requires only 16.03\,s using {\pkg}, a speed-up of 1.29. The exact speedup provided by parallel integration using {\pkg} depends on how independent the models are and how equally distributed the work is between the models. If the models are fully independent, the speedup is limited by the time required to execute the slowest model. If the models are dependent on one-another, the models may have to wait to receive messages and the speedup will not be as great. \S\ref{SS:speedup} provides additional information about the speedups achievable through {\pkg} parallel integration and how model structure influences the size of the performance boost.

While every model is executed on a new process, how the model is handled depends on the language it is written in. {\pkg} has a dedicated driver for each of the supported languages with utilities that allow {\pkg} to launch and monitor executables written in that language from Python. Models written in interpreted languages (Python and Matlab) are executed on the command line via the interpreter. In the case of Matlab, where a significant amount of time is required to start the Matlab interpreter (See \S\ref{SS:results_language}), Matlab shared engines are used to execute Matlab models. Matlab shared engines are Matlab instances that other processes can submit Matlab code to for execution. While shared engines also require the same amount of time to start as a standard instance of the Matlab interpreter, they can be started in advance and then reused.

@@

%---------------------------------------------------------------------------------
% COMMUNICATION
\subsection{Communication}\label{SS:communication}
%
In order for integrations to be possible, models must be able to send and receive information to and from other models written in different languages via some communication mechanism. Communication within {\pkg} integration networks was designed to be flexible in terms of the languages, platforms, and data types available. To accomplish this, {\pkg} leverages three different tools for communication, System V IPC Queues (\S\ref{SSS:ipc}), ZeroMQ (\S\ref{SSS:zmq}), and RabbitMQ (\S\ref{SSS:rmq}), which each have their own strengths and weaknesses. The particular communication mechanism used by {\pkg} is determined by the platform, available libraries, and integration strategy. However, regardless of the communication mechanisms, the user will always use the same interface in the language of their model (See \S\ref{SSS:interface}).

@@

%---------------------------------------------------------------------------------
% SYSTEM V IPC
\subsubsection{System V IPC Queues}\label{SSS:ipc}
%
The first communication mechanism used by {\pkg} was System V interprocess communication (IPC) message queues \citep{Rusling1999} on Posix (Linux and Mac OS X) systems. IPC message queues allow messages to be passed between models running on separate processes on the same machine. While IPC message queues are simple, fast (See \S\ref{SS:results_commtype}), and are built into most Posix operating systems, they do not work in all situations. IPC queues are not natively supported by Windows operating systems and do not allow communication between remote processes. In addition, IPC queues also have relatively low default message size limits on Mac OS X systems (2048 bytes). Once the queue is full or if the message is larger than the limit, any process attempting to send an additional message will block until a sufficient number of messages has been removed from the queue to accommodate the new message. For messages larger than the limit, the sending process will block indefinitely. This can be handled by splitting large messages into multiple smaller messages (See \S\ref{SSS:large}), however, the time required to send a message increases with the number of message it must be broken into (See \S\ref{SS:results_commtype}). These limits make sending large messages relatively inefficient when compared with other communications mechanisms. As a result, IPC queues are used by {\pkg} as a fallback on Posix systems if none of the other supported communication libraries have not been installed.

@@

%---------------------------------------------------------------------------------
% ASYNCHRONOUS MESSAGE MASSING
\subsubsection{Asynchronous Message Passing}\label{SSS:async}
%
The same asynchronous communication strategy is used with each of the communication mechanisms supported by {\pkg}. Models do not block on sending messages to output channels; the model is free to continue working on its task while an \emph{output driver} waits for the message to be routed and received on a separate master process thread. Similarly, an \emph{input driver} continuously pings input channels on another thread, moving received messages into a buffer queue so that they are ready and waiting for the receiving model when it asks for input. These drivers work together to move message along from one model to the next like a conveyor belt. As a result, models in complex integration networks are not affected by the rate at which dependent model consume their output and the speedup offered by running the models in parallel is improved. Figure \ref{fig:async} describes the general flow of messages, Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} in \A\ref{A:algorithms} describe the asynchronous output \& input channel procedures, and Algorithm \ref{alg:conn} describes the procedure followed by input \& output connection drivers.
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/io_drivers.png}
	\caption{Diagram of how messages are passed asynchronously using input/output drivers and an intermediate channel.}
	\label{fig:async}
	\end{center}
	\end{figure}
\fi
%
\begin{enumerate}
	\item A model sends a message in the form of a native data object via a language-specific API to one of the output channels declared in the model YAML. The output channel interface encodes the message and sends it.
	\item An output connection driver (written in Python) runs in a separate thread on the master process, listening to the model output channel. When the model sends a message, the output connection driver checks that the message is in the expected format and then forwards it to an intermediate channel. The intermediate channel is used as a buffer for support on distributed architectures (e.g. if one model is running on a remote machine). In these cases, the intermediate channel will connect to a RMQ broker using security credentials.
	\item An input connection driver (also written in Python) runs in a separate thread, listening to the intermediate channel. When a message is received, it is then forwarded to the input channel of the receiving model as specified in the integration network YAML.
	\item The receiving model receives the message in the form of an analogous native data object. Interface receive calls can be either blocking or non-blocking, but are blocking by default.
\end{enumerate}

@@

There are other optional model and input/output keywords for more advanced cases (e.g. compilation flags or data types), but they are beyond the scope of this limited introduction. Additional examples and descriptions of these options can be found in the documentation \footnote{{\tt https://cropsinsilico.github.io/cis\_interface/yaml.html}}.

@@

Once an integration pattern is selected, the wrappers can be flushed out in pseudocode independently by assuming that all input is received from a file and all output is sent to a file (i.e. the models are independent). The pseudocode for the root and shoot model wrappers is shown in Algorithms \ref{alg:root} and \ref{alg:shoot} respectively.
%
\begin{algorithm}[H]
    \caption{Root wrapper}
    \label{alg:root}
    \begin{algorithmic}[1] 
    \State \textbf{Receive} $r_r, R_t$
    \State \textbf{Send} $R_{t}$ \Comment{So that there is a complete record.}
    \While{true}
        \If {$dt$ available}
        		\State \textbf{Receive} $dt$
	\Else
		\State \textbf{Break}
	\EndIf
    	\State $R_{t+1} \gets {\tt calc\_root\_mass}(r_r, dt, R_t)$
	\State $R_t \gets R_{t+1}$
	\State \textbf{Send} $R_{t}$
    \EndWhile
    \end{algorithmic}
\end{algorithm}
%
\begin{algorithm}[H]
    \caption{Shoot wrapper}
    \label{alg:shoot}
    \begin{algorithmic}[1] 
    \State \textbf{Receive} $r_s, S_t, R_t$
    \State \textbf{Send} $S_{t}$ \Comment{So that there is a complete record.}
    \While{true}
        \If {$dt$ available}
        		\State \textbf{Receive} $dt$
	\Else
		\State \textbf{Break}
	\EndIf
	\State \textbf{Receive} $R_{t+1}$
	\State $S_{t+1} \gets {\tt calc\_shoot\_mass}(r_s, dt, S_t, R_t, R_{t+1})$
	\State $S_t \gets S_{t+1}$
	\State $R_t \gets R_{t+1}$
	\State \textbf{Send} $S_{t}$
    \EndWhile
    \end{algorithmic}
\end{algorithm}
%
Both models follow a similar pattern. They first receive static variables that will not change over the course of the run and then send the initial mass to output so that the output record is a complete history of the mass. Then both entire a while loop that is only broken when there is not a new time step available. When a new time step is available, it is received along with the next root mass in the case of the shoot model. Next, both model wrappers make the call to the actual model function. Finally, the models advance the time step by setting the masses to those calculated for the next time step and output the calculated mass.

@@

The first step in each model wrapper is to connect to the appropriate channels via the {\pkg} interface. This step occurs on lines 14-17 in the root model wrapper (Listing \ref{lst:wrap:root}) and lines 8-13 in the shoot model wrapper (Listing \ref{lst:wrap:shoot}). Regardless of the language, each has a similar call signature. Inputs require a single input that is a string specifying the name of the model input from the model YAML that the returned object should access. Outputs are similar in that their first argument is a string specifying the name of the model output from the mode YAML that the returned object should access. In addition, outputs can also be provided with a format string that tells {\pkg} interface what data type to expect how the output should be formatted if it is sent to a file. Additional information about how these strings are processed can be found in the ``C-Style Format Strings" section of the documentation\footnote{{\tt https://cropsinsilico.github.io/cis\_interface/c\_format\_strings.html}}.

@@

There are many file formats that {\pkg} supports which have additional YAML options.
Information about these file formats and their YAML options can be found in the ``Connection Options" subsection of the ``YAML Files" section of the documentation\footnote{{\tt https://cropsinsilico.github.io/cis\_interface/yaml.html\#connection-options}}.

@@

Except for Matlab models, the time required per message is not affected by the language of either model for smaller messages. There is a very small increase in the time per message when there is a Python model involved, but this is much lower than the standard deviation resulting from background activity on the test machine and so is in conclusive. Matlab models have a much higher time per message than any other language. This results from the way the Matlab interface was implemented, by calling the Python interface. The time required per message is largest when the receiving model is written in Matlab because the receiving models both receives messages from the other model and send output to a file, resulting in twice the number of calls to the wrapped interface. The performance of the Matlab interface may be improved in the future by either implementing it in Matlab directly or by wrapping the C routines.

@@

%---------------------------------------------------------------------------------
% COMMENTS
\newpage
\todo{To do:
\begin{itemize}
\item Replace links with those referencing new package name
\item Forking example
\item Transformation example
\item Add JSON to documentation
\item Add forking to documentation
\item Add transformation to documentation
\item Add schema to documentation
\item Regenerate documentation
\item Fix matlab autodocs for interface
\item Check that final version of algorithms is split between pages correctly
\end{itemize}
}

>>>>

%---------------------------------------------------------------------------------
% CONTROL THINGS
\newcommand{\todo}[1]{{\color{red}{#1}}}
\newcommand{\addref}{{\color{red}(ref. needed)}}
\newcommand{\Msol}{\,M$_{\odot}$}
\newcommand{\TB}{\,Tbytes}
\newcommand{\pkg}{{\tt yggdrasil}{}}
\newcommand{\cis}{{\tt cis\_interface}{}}
\newcommand{\pkglink}{\todo{cis\_interface}}
%\newcommand{\pkg}{{\tt \todo{cis\_interface}}{}}
\newcommand{\pkgrun}{{\tt \todo{cisrun}}}
\newcommand{\A}{Appendix{ }}
\newcommand{\hrefgit}[1]{\href{https://github.com/cropsinsilico/cis_interface}{\todo{#1}}}
\newcommand{\hrefdoc}[1]{\href{https://cropsinsilico.github.io/cis_interface/}{\todo{#1}}}
\newcommand{\hrefhack}[1]{\href{https://cropsinsilico.github.io/cis_interface/hackathon2018/index.html}{\todo{#1}}}

@@

%---------------------------------------------------------------------------------
% ABSTRACT (145 words)
\begin{abstract}
Thousands of computational models have been created within both the plant biology community and broader scientific communities in the past two decades that have the potential to be combined into complex integration networks capable of capturing more complex biological processes than possible with isolated models. However, the technological barriers introduced by differences in language and data formats has slowed this progress. We present \hrefgit{\pkg} (previously {\cis}), a Python package for running integration networks with connections between models across languages and scales. {\pkg} coordinates parallel execution of models in Python, C, C++, and Matlab on Linux, Mac OS, and Windows operating systems, and handles communication in a number of data formats common to computational plant modeling. {\pkg} is designed to be user-friendly and can be accessed at {\tt https://github.com/cropsinsilico/\pkglink}. Although originally developed for plant models, {\pkg} can be used to connect computational models from any domain.
\end{abstract}

@@

Consider the task of integrating a simple root growth model with a shoot growth model. Our example root growth model is written in C (see Listing \ref{lst:src:root}) and can be expressed as
\begin{equation}
R_{t+1} = R_{t} \times r_{r} \times dt + R_{t}.
\end{equation}
The inputs to the root growth model are
\begin{itemize}
	\item $R_{t}$: the root mass at time $t$ (g)
	\item $r_{r}$: the relative root growth rate (hr$^{-1}$)
	\item $dt$: the time step (hr)
\end{itemize}
and the output of the root growth model is $R_{t+1}$, the root mass at the next time step. 
%
The shoot growth model is written in Python (see Listing \ref{lst:src:shoot}) and can be expressed as
\begin{equation}
S_{t+1} = S_{t} \times r_{s} \times dt + S_{t} - (R_{t+1} - R_{t}).
\end{equation}
The inputs to the shoot model are 
\begin{itemize}
	\item $S_{t}$: the shoot mass at time $t$ (kg)
	\item $r_{s}$: the relative shoot growth rate (d$^{-1}$)
	\item $dt$: the time step (d)
	\item $R_{t}$: the root mass at time $t$ (kg)
	\item $R_{t+1}$: the root mass at time $t+1$ (kg)
\end{itemize}
and the output of the shoot growth model is $S_{t+1}$, the shoot mass at the next time step.
%
While it is possible to run both models in isolation if all of the appropriate input variables are provided, the value of $R_{t+1}$ calculated by the root growth model could be used as input to the shoot growth model. However, because the two models are written in different languages, the two models cannot be directly integrated. To integrate the models a scientist must either "manually" integrate them by running one model and then the other or translate one of the models into the language of the other so that it can be called directly. For a tightly coupled set of models such as these where one model depends directly on the result from the other at the current time step, manual integration is very inefficient for more than a handful of time steps even when done using a script. Although translating one or both of these simple toy models would be relatively straight forward, actual plant models are much more complex and could contain thousands of lines of code that would be time consuming to translate and result in unnecessary duplication of model algorithms.

We present an Open Source Python package, {\pkg}, for creating and running integration networks by connecting existing computational models written in different programming languages. {\pkg} was developed as part of the Crops \emph{in Silico} \citep{Marshall-Colon2017} initiative to build a complete crop \emph{in silico} from the level of the genes to the level of the field. {\pkg} is available on Github\footnote{{\tt https://github.com/cropsinsilico/\pkglink}} with full documentation\footnote{{\tt https://cropsinsilico.github.io/\pkglink/}} and can be installed from PyPI\footnote{{\tt https://pypi.org/project/\pkglink/}} via {\tt pip install {\pkg}} or conda-forge\footnote{{\tt https://github.com/conda-forge/\pkglink-feedstock}} via {\tt conda -c conda-forge install {\pkg}}. Although developed for the purpose of integrating plant biology models, {\pkg} can be applied to any situation requiring coordination between models in the supported languages. \S\ref{S:background} provides background information on existing efforts for cross-language model integration, \S\ref{S:methods} describes the methods used by the {\pkg} package to integrate models, \S\ref{S:example} provides a worked example of integrating the two model presented above, \S\ref{S:results} presents several tests demonstrating the performance of message passing between integrated models, and \S\ref{S:discuss} describes a few use cases for {\pkg}, 
summarizes the features and limitations of the {\pkg} package, and outlines areas of ongoing and future development.

@@

These three questions can be addressed through manual integration (e.g. running the root model, converting the output root mass into the units expected by the shoot model, and then running the shoot model). However, manual integration is 1) computationally inefficient when many iterations are necessary, such as during parameter searches or steady state convergence tests, 2) unique to every integration, requiring the production of new scripts, 3) time consuming, 4) prone to error, and 5) complex when more than three models are being integrated. Within both computational biology and the larger scientific computing community, groups have developed around different solutions to the problem of connecting models. These solutions can generally be grouped into two categories: domain specific languages and frameworks.

@@

%---------------------------------------------------------------------------------
% ORCHESTRATION
\subsection{Orchestration}\label{SS:orchestration}
%
{\pkg} is executed via a command line interface (CLI), {\pkgrun} with specification files (\S\ref{SSS:yaml}) as input. Based on the information contained in the specification files, {\pkg} dynamically establishes a network of asynchronous communication channels (\S\ref{SS:communication}), and launches the models on new processes (\S\ref{SS:model_execution}). Although written in Python, {\pkg} was written such that users need not have any knowledge of the Python language and can interact with {\pkg} solely through the CLI. In addition to the CLI, more advanced capabilities are also exposed via the Python API including access to more detailed information about the running integration.
%At execution, {\pkg} launches models on new processes based on information contained in a YAML specification file (See \S\ref{SSS:yaml} below) and establishes

@@

Model YAMLs include information about the location of the model source code, the language the model is written in, how the model should be run, and any input or output variables including their data type (e.g. array, scalar, mesh) and physical units. 
%
%For example, Listings \ref{lst:yml:root} and \ref{lst:yml:shoot} from \A\ref{A:yml} are the model YAMLs for the example root and shoot growth models from \S\ref{S:intro}. These model YAMLs could be used in any integration without being modified because the model inputs and outputs will not change.
%
%\rootyml
%
%\shootyml
%
Integration networks are specified by declaring the connections between models and connections are declared by pairing an output variable from one model with the input variable of another model. The {\pkg} CLI sets up the necessary communication mechanisms to then direct data from one model to the next in the specified pattern. Models can have as many input and/or output variables as is desired and connections between models are specified by references to the input/output variables associated with each model. In addition, input and output variables can also be connected to files.
%
%For example, in the connection YAML for integrating the root and shoot models (Listing \ref{lst:yml:int} from \A\ref{A:yml}), the first two connections link the {\tt root\_input} and {\tt shoot\_input} input channels from the root and shoot model YAML's to input files. The third connection links the {\tt next\_root\_mass} output variable from the root model YAML (Listing \ref{lst:yml:root}) to the {\tt next\_root\_mass} input variable from the shoot model YAML (Listing \ref{lst:yml:shoot}). The forth connection links the {\tt next\_shoot\_mass} to an output file.
%
%\intyml
%
This format allows users the flexibility to create complex integration networks and test models in isolation before running the entire integration network as will be seen in \S\ref{S:example}.

%---------------------------------------------------------------------------------
% MODEL EXECUTION
\subsubsection{Model Execution}\label{SS:model_execution}
%
{\pkg} launches each model in an integration in its own process, allowing models to complete independent operations in parallel and complete tasks more quickly. For example, the root and shoot growth models from \S\ref{S:intro} (described further in \S\ref{S:example}) require 10.28\,s and 10.40\,s respectively to run for 100 time steps in isolation with direct input/output from/to files in their native language. If these two models were manually integrated in serial via the command line, the integration would require a total of 20.68\,s. However, because {\pkg} offers parallel execution of the models, the same integration requires only 16.03\,s when using {\pkg}, a speed-up of 1.29. The exact speedup provided by parallel integration using {\pkg} depends on how independent the models are and how equally distributed the work is between the models. If the models are fully independent, the speedup is limited by the time required to execute the slowest model. If the models are dependent on one-another, the models may have to wait to receive messages and the speedup will not be as great. \S\ref{SS:speedup} provides additional information about the speedups achievable through {\pkg} parallel integration and how model structure influences the size of the performance boost.

While every model is executed in a new process, how the model is handled depends on the language it is written in. {\pkg} has a dedicated driver for each of the supported languages with utilities that allow {\pkg} to launch and monitor executables written in that language from Python. Models written in interpreted languages (Python and Matlab) are executed on the command line via the interpreter. In the case of Matlab, where a significant amount of time is required to start the Matlab interpreter (see \S\ref{SS:results_language}), Matlab shared engines are used to execute Matlab models. Matlab shared engines are Matlab instances that other processes can submit Matlab code to for execution. While shared engines also require the same amount of time to start as a standard instance of the Matlab interpreter, they can be started in advance and then reused.

@@

%---------------------------------------------------------------------------------
% COMMUNICATION
\subsection{Communication}\label{SS:communication}
%
Integrating models requires that they are able to send and receive information to and from other models written in different languages via some communication mechanism. Communication within {\pkg} integration networks was designed to be flexible in terms of the languages, platforms, and data types available. To accomplish this, {\pkg} leverages three different tools for communication, System V IPC Queues (\S\ref{SSS:ipc}), ZeroMQ (\S\ref{SSS:zmq}), and RabbitMQ (\S\ref{SSS:rmq}), which each have their own strengths and weaknesses. The particular communication mechanism used by {\pkg} is determined by the platform, available libraries, and integration strategy. However, regardless of the communication mechanisms, the user will always use the same interface in the language of their model, simplifying the number of routines that users need to use for integration (see \S\ref{SSS:interface}).

@@

%---------------------------------------------------------------------------------
% SYSTEM V IPC
\subsubsection{System V IPC Queues}\label{SSS:ipc}
%
The first communication mechanism used by {\pkg} was System V interprocess communication (IPC) message queues \citep{Rusling1999} on Posix (Linux and Mac OS X) systems. IPC message queues allow messages to be passed between models running on separate processes on the same machine. While IPC message queues are simple, fast (See \S\ref{SS:results_commtype}), and are built into most Posix operating systems, they do not work in all situations. IPC queues are not natively supported by Windows operating systems and do not allow communication between remote processes. In addition, IPC queues also have relatively low default message size limits on Mac OS X systems (2048 bytes or 256 64 bit numbers). Once the queue is full or if the message is larger than the limit, any process attempting to send an additional message will stop until a sufficient number of messages has been removed from the queue to accommodate the new message. For messages larger than the limit, the sending process will stop indefinitely. This can be handled by splitting large messages into multiple smaller messages (see \S\ref{SSS:large}); however, the time required to send a message increases with the number of message it must be broken into (See \S\ref{SS:results_commtype}). These limits make sending large messages relatively inefficient when compared with other communications mechanisms. As a result, IPC queues are used by {\pkg} only as a fallback on Posix systems if none of the other supported communication libraries have not been installed.

@@

%---------------------------------------------------------------------------------
% ASYNCHRONOUS MESSAGE MASSING
\subsubsection{Asynchronous Message Passing}\label{SSS:async}
%
The same asynchronous communication strategy is used with each of the communication mechanisms supported by {\pkg}. Models do not block on sending messages to output channels; the model is free to continue working on its task while an \emph{output driver} waits for the message to be routed and received on a separate master process thread. Similarly, an \emph{input driver} continuously checks input channels on another thread, moving received messages into a intermediate buffer queue so that they are ready and waiting for the receiving model when it asks for input. These drivers work together to move message along from one model to the next like a conveyor belt. As a result, models in complex integration networks are not affected by the rate at which dependent model consume their output and the speedup offered by running the models in parallel is improved. Figure \ref{fig:async} describes the general flow of messages, Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} in \A\ref{A:algorithms} describe the asynchronous output \& input channel procedures, and Algorithm \ref{alg:conn} describes the procedure followed by input \& output connection drivers.
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/io_drivers.png}
	\caption{Diagram of how messages are passed asynchronously using input/output drivers and an intermediate channel.}
	\label{fig:async}
	\end{center}
	\end{figure}
\fi
%
\begin{enumerate}
	\item A model sends a message in the form of a native data object via a language-specific API to one of the output channels declared in the model YAML. The output channel interface encodes the message and sends it.
	\item An output connection driver (written in Python) runs in a separate thread on the master process, listening to the model output channel. When the model sends a message, the output connection driver checks that the message is in the expected format and then forwards it to an intermediate channel. The intermediate channel is used as a buffer for support on distributed architectures (e.g. if one model is running on a remote machine). In these cases, the intermediate channel will connect to a RMQ broker using security credentials.
	\item An input connection driver (also written in Python) runs in a separate thread, listening to the intermediate channel. When a message is received, it is then forwarded to the input channel of the receiving model as specified in the integration network YAML.
	\item The receiving model receives the message in the form of an analogous native data object. Interface receive calls can be either blocking or non-blocking, but are blocking by default.
\end{enumerate}

@@

There are other optional model and input/output keywords for more advanced cases (e.g. compilation flags or data types), but they are beyond the scope of this limited introduction. Additional examples and descriptions of these options can be found in the documentation \footnote{{\tt https://cropsinsilico.github.io/\pkglink/yaml.html}}.

@@

Once an integration pattern is selected, the wrappers can be flushed out in pseudocode independently by assuming that all input is received from a file and all output is sent to a file (i.e. the models are independent). The pseudocode for the root and shoot model wrappers is shown in Algorithms \ref{alg:root} and \ref{alg:shoot} respectively.
%
\begin{algorithm}[H]
    \caption{Root wrapper}
    \label{alg:root}
    \begin{algorithmic}[1] 
    \State \textbf{Receive} $r_r, R_t$
    \State \textbf{Send} $R_{t}$ \Comment{So that there is a complete record.}
    \While{true}
        \If {$dt$ available}
        		\State \textbf{Receive} $dt$
	\Else
		\State \textbf{Break}
	\EndIf
    	\State $R_{t+1} \gets {\tt calc\_root\_mass}(r_r, dt, R_t)$
	\State $R_t \gets R_{t+1}$
	\State \textbf{Send} $R_{t}$
    \EndWhile
    \end{algorithmic}
\end{algorithm}
%
\begin{algorithm}[H]
    \caption{Shoot wrapper}
    \label{alg:shoot}
    \begin{algorithmic}[1] 
    \State \textbf{Receive} $r_s, S_t, R_t$
    \State \textbf{Send} $S_{t}$ \Comment{So that there is a complete record.}
    \While{true}
        \If {$dt$ available}
        		\State \textbf{Receive} $dt$
	\Else
		\State \textbf{Break}
	\EndIf
	\State \textbf{Receive} $R_{t+1}$
	\State $S_{t+1} \gets {\tt calc\_shoot\_mass}(r_s, dt, S_t, R_t, R_{t+1})$
	\State $S_t \gets S_{t+1}$
	\State $R_t \gets R_{t+1}$
	\State \textbf{Send} $S_{t}$
    \EndWhile
    \end{algorithmic}
\end{algorithm}
%
Both models follow a similar pattern. They first receive ``static" variables that will not change over the course of the run and then send the initial mass to output so that the output record is a complete history of the mass. Then both entire a while loop that is only broken when there is not a new time step available. When a new time step is available, it is received along with the next root mass in the case of the shoot model. Next, both model wrappers make the call to the actual model function. Finally, the models advance the time step by setting the masses to those calculated for the next time step and output the calculated mass.

@@

The first step in each model wrapper is to connect to the appropriate channels via the {\pkg} interface. This step occurs on lines 14-17 in the root model wrapper (Listing \ref{lst:wrap:root}) and lines 8-13 in the shoot model wrapper (Listing \ref{lst:wrap:shoot}). Regardless of the language, each has a similar call signature. Inputs require a single input that is a string specifying the name of the model input from the model YAML that the returned object should access. Outputs are similar in that their first argument is a string specifying the name of the model output from the mode YAML that the returned object should access. In addition, outputs can also be provided with a format string that tells {\pkg} interface what data type to expect how the output should be formatted if it is sent to a file. Additional information about how these strings are processed can be found in the ``C-Style Format Strings" section of the documentation\footnote{{\tt https://cropsinsilico.github.io/\pkglink/c\_format\_strings.html}}.

@@

There are many file formats that {\pkg} supports which have additional YAML options.
Information about these file formats and their YAML options can be found in the ``Connection Options" subsection of the ``YAML Files" section of the documentation\footnote{{\tt https://cropsinsilico.github.io/\pkglink/yaml.html\#connection-options}}.

@@

Except for Matlab models, the time required per message is not affected by the language of either model for smaller messages. There is a very small increase in the time per message when there is a Python model involved, but this is much lower than the standard deviation resulting from background activity on the test machine and so is inconclusive. Matlab models have a much higher time per message than any other language. This results from the way the Matlab interface was implemented, by calling the Python interface. The time required per message is largest when the receiving model is written in Matlab because the receiving models both receives messages from the other model and send output to a file, resulting in twice the number of calls to the wrapped interface. The performance of the Matlab interface may be improved in the future by either implementing it in Matlab directly or by wrapping the C routines.

@@

%---------------------------------------------------------------------------------
% COMMENTS
%\newpage
%\todo{To do:
%\begin{itemize}
%\item Replace links with those referencing new package name
%\item Forking example
%\item Transformation example
%\item Add JSON to documentation
%\item Add forking to documentation
%\item Add transformation to documentation
%\item Add schema to documentation
%\item Regenerate documentation
%\item Fix matlab autodocs for interface
%\item Check that final version of algorithms is split between pages correctly
%\end{itemize}
%}
